What we {\sf eval} in the shadows

### 2021-04-01 paper/main.tex
=> 2.3

\citet{ecoop12} gave a programming language-centric overview of the R language.
They characterized it as a lazy, vectorized, functional language with rich
dynamic features that allow layering several object systems. Most data types are
sequences of primitive values constructed by calling the combination function
\c{c}, so that \c{c(1,2)} evaluates to a vector of two strings, and constants
such as \c{42} are vectors of length one. To enable equational reasoning, values
accessible through multiple aliases are copied when written to. Furthermore,
values can be tagged by user-defined attributes, key-value pairs. For instance,
the attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the
call \c{attr(x,"dim")<-c(2,2)}. In this case, the addition of this attribute
ensure that the vector \c x will be treated as a $2 \times 2$ matrix by all
arithmetic operations. The \c{class} attribute is used to give a 'class', in the
object-oriented sense, to values. So, \c{class(x)<-"human"} sets the class of
\c{x} to \c{human}; classes are used for method dispatch. Every linguistic
construct is desugared to a function call, even control flow statements,
assignments, and bracketing. All functions can be shadowed or redefined, making
the language at the same time remarkably flexible and exceeding challenging to
compile statically as vividly detailed by~\citet{dls19}. R uses a relaxed call-by-need convention for
passing arguments to functions. Each argument is a thunk composed of an
expression, its environment, and a slot for the result, these are called
\emph{promises}. To get the value of an argument, the corresponding promise must
be forced, once forced the promise's result is cached for future use.

\subsection{The Semantics of Eval and Friends}

@@

\begin{figure}[!t]\center\small
\begin{tabular}{r|l|l|l}\hline
\tiny\sc Language&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
\bf Julia     & expression& toplevel         & ??\\
\bf Java      & bytecode  & classloader       & data\\
\bf JavaScript& text      & current, toplevel& data\\
\bf R         & expression& programmatic      & data, stack, envir.\\\hline
\end{tabular}
\caption{Design space of \eval}\label{comp}
\end{figure}

@@

WORK IN  PROGREESS

JavaScript and R evaluate it in the current environment, thus exposing local
variables and parameters and breaking the caller's abstraction boundary. Julia
is more restrictive, \eval runs at the ``top level'', in the global environment.
JavaScript has a global \eval that behaves like Julia, and a ``strict mode'' in
which \eval may access local variables, but fobidden from injecting new ones in
the local environment. In Java, one can implement \eval with
Julia-semantics.\footnote{While \eval used to be the purview of interpreted
languages, just-in-time compilation lifted this restriction. \Eval takes its
input, wraps it in a static method of a new anonymous class, generates bytecode
for that class, invokes the class loader to install that code, and finally
reflectively calls the method. }

 {\bf Reflective API:} What other reflective operations are
  exposed to user code? JavaScript provides few reflective functions
  (other than enumeration of an object's properties and string-based
  indexing), the impact of \eval is thus limited. R, on the other
  hand, has a rich reflective API. For example, R allows user code to
  walk the call stack and arbitrarily add and delete local variables.
  This means that there are few limits to the side-effects of an
  \eval.

\noindent
These design choices impact our ability to reason about code when
reflection is used. A more expressive \eval translates directly into
additional restrictions for program analysis and transformation tools.
The reason Julia restricts \eval to the top level is to preserve the
compiler's ability to optimize code~\cite{oopsla18a}. In particular,
\eval had to be prevented from changing the type of local variables as
this would require recompilation of the method. Julia even adds a
versoning mechanism, called world age, to ensure that code added
during an \eval does not invalidate inlining
optimizations~\cite{oopsla20a}. More permissive languages such as R
are much harder to deal with from a compiler's perspective.

\mypara{Eval in R.} The reflective interface exposed by
R is somewhat complex. The core library exports four functions with
slightly different semantics, \c{eval}, \c{evalq}, \c{eval.parent},
and \c{local}. \Eval is the more general function, and, unlike in
JavaScript, it takes three arguments, an expression, its environment
and the enclosing environment. The following discussion simplify some
details which are not relevant to this paper. For the interested
reader we recommend the excellent book by Hadley
Wickham~\cite{hadley}. The definition of \eval starts as follows:

\begin{lstlisting}
 eval <- function(expr, envir = parent.frame(),
           enclos = if(is.list(envir)) parent.frame() else baseenv()) {...body...}
\end{lstlisting}
The parameters are \c{expr}, the value to be evaluated, \c{envir}, the
environment in which evaluation happens, and \c{enclos}, the
environment to look up objects not found in \c{envir} (by default
either the top level, \c{baseenv} or the environment of the caller).
The \c{expr} parameter can take values of different type, for our
purposes we focus on the most common, {\tt expression}s. It is easiest
to think of an \c{expression} as an abstract syntax tree as returned
by R's parser. The simplest way to create an \c{expression} is to call
\c{quote} passing some R expression:
\begin{lstlisting}
  > quote(a + b)
  # a + b
\end{lstlisting}
Here the return value is an abstract syntax tree. From a
string, use  the \c{parse} function:
\begin{lstlisting}
  > parse(text="a+b")
  # expression(a+b)
\end{lstlisting}
The most common way to create an expression is to extract it from an
argument. Each argument comes packaged inside a promise which retains
the source code of the argument. Consider the following function
definition, \c{f} is a function with a single parameter \c{x}. When
the call \c{f(a+b)} is evaluated, \c{a+b} is stored inside a newly
created promise.
\begin{lstlisting}
  > f <- function(x) substitute(x), list(a=1)))
  > f(a+b)
  # expression(1+b)
  > substitute(a+b,list(a=1,b=2))
  # 1+2
\end{lstlisting}
The call to \c{substitute(x)} extracts, from the promise bound to
\c{x}, the expression passed in the call to \c{f}, i.e., \c{a+b}. The
\c{substitute(exp,env)} function takes two arguments, the second is an
object that can be used as an environment. This can be either a list
of named values, a data frame or an actual environment. \c{substitute}
will look for occurrences of each symbol from \c{env} in \c{exp} and
replace them with their value taken from \c{env}.
\begin{lstlisting}
  > environment()
  # <environment: R_GlobalEnv>
\end{lstlisting}
The above shows how to access the current environment, this can belong
to the current function or, as above, the top level. Environments are
nested, with each environment having a parent. This nesting is used
when looking up names. When calling \c{new.env}, the default parent is
the current environment. The chain of environments can be traversed
with \c{parent.env}, ultimately one arrives at \c{emptyenv}.
\c{parent.frame} and \c{sys.frame} grant access to environments
further up in the calling stack, up to the global environment
\c{.GlobalEnv}, and packages. One can also directly read, modify or
create new bindings, given any environment:
\begin{itemize}
\item \c{env\$v} and \c{get("v",envir=env)} read variable \c{v} in
  environment \c{env};
\item \c{env\$v<-2} and \c{assign("v",2,envir=env)} write 2 in
  \c{v}. If not found, \c{v} is created in \c{env}.
\end{itemize}
\noindent Environments are often used as hashtables by programmers as
they have reference semantics (other are copy-on-write) and have a
built-in string look up. across function calls, or to create package
namespaces.
\begin{lstlisting}
  evalq <- function(exp,env,enclos) eval(quote(exp),env,enclos)
  eval.parent <- function(exp,n) eval(exp,parent.frame(n))
  local <- function(exp,env,enclos) evalq(exp,new.env())
\end{lstlisting}
The three \eval variants can be expressed as calls to \eval. The
\c{evalq} form quotes the argument to prevent it from being evaluated
in the current environment. The \c{eval.parent(e,n)} form evaluates
\c{e} in the environment of the {\tt n}-th caller of this function.
Finally, \c{local} evaluates an \c{exp} in a new environment to
avoid polluting the current one.

>>>>

\citet{ecoop12} gave a programming language-centric overview of the R language.
They characterized it as a lazy, vectorized, functional language with rich
dynamic features that allow layering several object systems. Most data types are
sequences of primitive values constructed by calling the combination function
\c{c}, so that \c{c(1,2)} evaluates to a vector of two strings, and constants
such as \c{42} are vectors of length one. To enable equational reasoning, values
accessible through multiple aliases are copied when written to. Furthermore,
values can be tagged by user-defined attributes, key-value pairs. For instance,
the attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the
call \c{attr(x,"dim")$\leftarrow$c(2,2)}. In this case, the addition of this
attribute ensure that the vector \c x will be treated as a $2 \times 2$ matrix
by all arithmetic operations. The \c{class} attribute is used to give a 'class',
in the object-oriented sense, to values. So, \c{class(x)<-"human"} sets the
class of \c{x} to \c{human}; classes are used for method dispatch. Every
linguistic construct is desugared to a function call, even control flow
statements, assignments, and bracketing. All functions can be shadowed or
redefined, making the language at the same time remarkably flexible and
exceeding challenging to compile statically as vividly detailed
by~\citet{dls19}. R uses a relaxed call-by-need convention for passing arguments
to functions. Each argument is a thunk composed of an expression, its
environment, and a slot for the result, these are called \emph{promises}. To get
the value of an argument, the corresponding promise must be forced, once forced
the promise's result is cached for future use.

\subsection{On the Expressive Power of Eval}

@@

\begin{figure}[!t]\center\small
\begin{tabular}{r@{~}l|l|l|l}\hline
\tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
\bf Julia&\cite{julia}     & expression& toplevel         & data\\
\bf Java&\cite{cl}  & bytecode  & classloader       & data\\
\bf JavaScript&\cite{ecoop11}& text      & current, toplevel& data\\
\bf R&\cite{R96}  & expression& programmatic      & data, stack, envir.\\\hline
\end{tabular}
\caption{Design space of \eval}\label{comp}
\end{figure}

@@

The last degree of freedom in the design of \eval is what can be done from
within the generated code. The main difference between designs lies in what part
of the system state can be accessed and modified. In R, there is very little
that cannot be observed from an \eval context and modified. The code can
override the definition of variables and functions. Also, thanks to a rich
reflective interface, code can introspect on values, walk and modify environment
chains, and access the stack frames on the call stacks and their environments.

Given the above, it is reasonable to claim that R's \eval are amongst the
languages that afford the facility the most expressive power. This comes at the
cost of the programmer's ability to reason about code in the presence of \eval.
Expressive power translates directly into limits on the usefulness of program
analysis.

By contrast, the designers of Julia chose to limit \eval. In Julia, only global
variables can be side-effected and environments cannot be readily manipulated.
This is designed to shield optimized code from some of the most pernicious uses
of the facility~\cite{oopsla18a}. Furthermore, the language provides a
versioning mechanism, called world age, to ensure that any methods defined
within an \eval only become visible at well-defined program points and thus that
recompilation does not have to occur when optimized code is
running~\cite{oopsla20a}.

\subsection{Eval in R}

The R language exposes a rich reflective interface with four functions with
different semantics, \c{eval}, \c{evalq}, \c{eval.parent}, and
\c{local}. The first is the most general, taking three arguments, an expression,
its environment and an enclosing environment. The following discussion
simplifies details not relevant to this paper, the interested reader should
consult \citet{hadley}. It is defined as:

\begin{lstlisting}
 eval <- function(e, envir = parent.frame(),
                   encl = if(is.list(envir)) parent.frame() else baseenv()) ...
\end{lstlisting}
Parameter \c e is the expression to evaluate, \c{envir} is the evaluation
environment, and \c{encl} is used to look up variables not found in \c{envir}.
The latter is, by default, either the environment of the function that called
\eval or the top level. Parameter \c e can be of different types, for our
purposes we focus on {\tt expression}s. These can be thought of as abstract
syntax trees returned by the parser. An \c{expression} can be obtained by
calling \c{quote} with an expression or \c{parse} wit a string:
\begin{lstlisting}
 > quote(a + b)
 # a + b
 > parse(text="a+b")
 # expression(a+b)
\end{lstlisting}
Even more commonly, expressions are obtained from promises using \c{substitute}.
Consider a one-parameter function \c{f}, when called with \c{a+b}, the function
returns \c{a+b} as expression.
\begin{lstlisting}
 > f <- function(x) substitute(x)
 > f(a+b)
 # expression(a+b)
 > substitute(a+b,list(a=1,b=2))
 # 1+2
\end{lstlisting}
The call to \c{substitute(x)} extracts the unevaluated expression from the
promise bound to variable \c x. Function \c{substitute(e,envir)} takes two
arguments, the second is used as a substitution environment for free variables
in \c e.
\begin{lstlisting}
 > environment()
 # <environment: R_GlobalEnv>
\end{lstlisting}
The \c{environment} function returns a variable environment, either from the
currently executing function or the top level environment. Environments nest,
each has a parent. When creating a new environment with \c{new.env}, the parent
is the current environment. Environment chains can be traversed with
\c{parent.env}, until \c{emptyenv} is reached. The top level environment is
\c{.GlobalEnv}, it has parents that represent the packages that have been
loaded. One can also directly read, modify or create new bindings, given any
environment:
\begin{itemize}
\item \c{envir\$v} and \c{get("v",envir=envir)}: read  \c{v} from \c{envir};
\item \c{envir\$v<-2} and \c{assign("v",2,envir=envir)}: store 2 in \c{v}, if
  not found, \c{v} is created in \c{envir}.
\end{itemize}
\noindent Programmers sometimes use environments as hash maps as they are the
only data type with reference semantics and have a built-in string look up.
Functions \c{parent.frame} and \c{sys.frame} return environments further up the
call stack.
 \begin{lstlisting}
  evalq <- function(e,envir,encl) eval(quote(e),envir,encl)
  eval.parent <- function(e,n) eval(e,parent.frame(n))
  local <- function(e,envir,encl) evalq(e,new.env())
\end{lstlisting}
The three \eval variants can be expressed with \eval. The \c{evalq} form quotes
the argument to prevent it from being evaluated in the current environment. The
\c{eval.parent(e,n)} form evaluates \c{e} in the environment of the {\tt n}-th
caller of this function. Finally, \c{local} evaluates \c{e} in a new environment
to avoid polluting the current one.
